---
title: "STAT 6348 Applied Multivariate Analysis Project 1"
author: "John Kenney"
date: "2024-02-16"
output: github_document
header-includes: 
  - \usepackage{fvextra}
  - \usepackage{float} 
  - \floatplacement{figure}{H}
  - \usepackage{bm}
  - \usepackage{wrapfig}
  - \usepackage{amsmath}
  - \usepackage{mathtools}
  - \usepackage{longtable}
  - \usepackage{enumitem}
  - \usepackage{booktabs}
  - \usepackage{verbatim}
  - \allowdisplaybreaks
  - \usepackage{fullpage}
  - \usepackage{subfig}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DeclareUnicodeCharacter{2212}{-}
urlcolor: blue

always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message = FALSE,warning = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```





```{r,echo=F, results='hide',fig.show='hide'}
#HCV <- read.csv("https://utdallas.box.com/shared/static/6noqrx35iao10p4w15z7kx6egggiiq8f.csv")
HCV <- read.csv("Project_1_multivariate_analysis_and_assumptions/Data/HCV.csv")
HCV$Category <- factor(HCV$Category)
tmp <- HCV
HCV$Sex <- factor(HCV$Sex)
```
**Question 1**    
  
Consider the HCV data. The data set contains laboratory values of blood donors and Hepatitis C patients (including its progress - 'just' Hepatitis C, Fibrosis, Cirrhosis) and demographic values age and sex. The variables in the dataset are the following:    
   
The laboratory data are the attributes/variables 5-14.   
   
1)  X (Patient ID/No.)    
2)  Category (diagnosis): '0=Blood Donor', '1=Hepatitis', '2=Fibrosis', '3=Cirrhosis'   
3)  Age (in years)   
4)  Sex (f, m)   
5)  ALB   
6)  ALP   
7)  ALT   
8)  AST   
9)  BIL   
10) CHE   
11) CHOL   
12) CREA   
13) GGT   
14) PROT   
   


```{r,echo=F, results='hide',fig.show='hide'}
HCV_num <- subset(HCV,select =-c(Category,Age,Sex))
#head(HCV_num)

```
 
**Question 1a)**    
  
Calculate Mahalanobis distances of each observation vector (with laboratory variables only) to the center in terms of the covariance matrix. Report only the summary statistics of the distances. Which patients are nearest and furthest to the center and how do they compare? How would these results change if Euclidean distance was used instead?  

**Question 1a Answer)**   
```{r,echo=F, results='hide',fig.show='hide'}
#Calculate Mahalanobis Distances
maha.dist<- sqrt(mahalanobis(HCV_num,center=colMeans(HCV_num),cov=cov(HCV_num)))
names(maha.dist) <- paste("Patient_",1:(dim(HCV_num)[1]),sep="")
#Sort the Distances
#sort(maha.dist)

# mahalanobis distance summary
maha.summary <- summary(maha.dist)
dig <- 4
table.maha.summary <- as.data.frame(
  cbind(round(maha.summary[[1]],dig), round(maha.summary[[2]],dig), round(maha.summary[[3]],dig), 
        round(maha.summary[[4]],dig), round(maha.summary[[5]],dig), round(maha.summary[[6]],dig)))
colnames(table.maha.summary) <- c("Min","1st.Quantile","Median","Mean","3rd.Quantile","Max")
table.maha.summary <- rbind(table.maha.summary,
                            c(names(which(maha.dist == min(maha.dist))), "", names(which(maha.dist == median(maha.dist))), "", "", names(which(maha.dist == max(maha.dist)))))

# kable extratable creation
library(kableExtra)
knitr::kable(table.maha.summary,booktabs = T, 
             caption = "Summary Statistics of Mahalanobis distances", 
             align = "cccccc") %>% 
  kable_styling(latex_options = c("hold_position"))
```


Patient 325 is the closest to the center based on mahalanobis distance and Patient 570 is the furthest away from the center. We can see that Patient 325's distance is 0.7105 and Patient 570's distance is 19.8253 therefore there seems a large difference between these two patients. If we look at both patients data Patient 325 is a Blood Donor, 33 years old, and female and Patient 570 has a diagnosis of Cirrhosis, 46 years old, and Male, so it makes sense that their lab work would be far apart.

```{r,echo=F, results='hide',fig.show='hide'}
# to look at patient 325 and 570
#HCV[c(325,570),]

library(gridExtra)

```


```{r,echo=F, results='hide',fig.show='hide'}
# euclidean function
euclidean <- function(x) {
  dist <- sqrt(rowSums(x -  colMeans(x))^2)
  return(dist)
}

#Calculate Mahalanobis Distances
euclidean.dist <- euclidean(HCV_num)
names(euclidean.dist) <- paste("Patient_",1:(dim(HCV_num)[1]),sep="")
#Sort the Distances
#sort(euclidean.dist)

# euclidean distance summary
euclidean.summary <- summary(euclidean.dist)
dig <- 4
table.euclidean.summary <- as.data.frame(
  cbind(round(euclidean.summary[[1]],dig), round(euclidean.summary[[2]],dig), round(euclidean.summary[[3]],dig), 
        round(euclidean.summary[[4]],dig), round(euclidean.summary[[5]],dig), round(euclidean.summary[[6]],dig)))
colnames(table.euclidean.summary) <- c("Min","1st.Quantile","Median","Mean","3rd.Quantile","Max")
table.euclidean.summary <- rbind(table.euclidean.summary,
                            c(names(which(euclidean.dist == min(euclidean.dist))), "", names(which(euclidean.dist == median(euclidean.dist))), "", "", names(which(euclidean.dist == max(euclidean.dist)))))

# kable extratable creation
library(kableExtra)
knitr::kable(table.euclidean.summary,booktabs = T, 
             caption = "Summary Statistics of Euclidean distances", 
             align = "cccccc") %>% 
  kable_styling(latex_options = c("hold_position"))
```


The patient closest to the center based on Euclidean distance is Patient 465 and the patient furthest away from the center Patient 570 did not change with the different distance metric used. 
   
   
   
   
**Question 1b)**     
   
What are the directions and half-lengths of the two longest axes of the ellipsoid?
   
**Question 1b Answer)**   
$(X - \overline{X})^\top \Sigma^{-1}(X - \overline{X}) \le c^2$    
   
The half-length of the largest axis (major) is $c\sqrt{\lambda_1} = c\times\sqrt{3590.537} = c \times 59.92109$ from the origin $\overline{X}$ in the direction of the vector $e_1 = [0.015621802, -0.2240444617, -0.0846236613, -0.3032719000, -0.0710508473,$  
$0.0039750179,  0.0005335915, -0.3245507772, -0.8602402464,  0.0037725469]$  
 
If large sample theory holds we can use $c = \sqrt{\chi^2_p(\alpha)} = \sqrt{\chi^2_{10}(0.95)} = \sqrt{18.30704} = 4.278673$. Then the largest half-length would be 256.3827.

The half-length of the second largest axis (minor) is $c\sqrt{\lambda_2} = c\times\sqrt{2476.81} = c \times 49.76756$ from the origin $\overline{X}$ in the direction of the vector $e_2 = [0.0081947233, -0.0044550145, -0.0637826732, -0.1980907922,$   
$-0.0326184499,  0.0017948186, -0.0006255074,  0.9383531779, -0.2739091658, -0.0020771672$   
 
If large sample theory holds we can use $c = \sqrt{\chi^2_p(\alpha)} = \sqrt{\chi^2_{10}(0.95)} = \sqrt{18.30704} = 4.278673$. Then the second largest half-length would be 212.9391.



```{r,echo=F, results='hide',fig.show='hide'}
# covariance matrix

# X <- as.matrix(HCV_num)
# (t(X) %*% (diag(dim(X)[1]) - (1/dim(X)[1])*matrix(1,nrow = dim(X)[1],ncol = dim(X)[1])) %*% X)*(1/(dim(X)[1]-1))
S <- cov(HCV_num,method = "pearson")

# get two largest eigen pairs of of the ellipsoid
#eigen(S)
lam1 <- eigen(S)$values[1]
e1 <- eigen(S)$vectors[,1]
lam2 <- eigen(S)$values[2]
e2 <- eigen(S)$vectors[,2]

```



**Question 1c)**    
  
For laboratory variables, calculate sample mean vector, covariance matrix, and correlation matrix. Create a correlation heatmap in which intensity of colors reflects the strength of correlation. Add a legend to the heatmap.  

**Question 1c Answer)**   
Refer to the output section for Question 1c) answer.
```{r,echo=F, results='hide',fig.show='hide'}
# sample mean vector
xbar <- colMeans(HCV_num)

dig <- 4

# kable extratable creation
library(kableExtra)
knitr::kable(round(t(xbar),dig),booktabs = T, 
             caption = "Sample Mean Vector of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position"))
```


```{r,echo=F, results='hide',fig.show='hide'}
# covariance matrix

# X <- as.matrix(HCV_num)
# (t(X) %*% (diag(dim(X)[1]) - (1/dim(X)[1])*matrix(1,nrow = dim(X)[1],ncol = dim(X)[1])) %*% X)*(1/(dim(X)[1]-1))
S <- cov(HCV_num,method = "pearson")



dig <- 3

# kable extra table creation
library(kableExtra)
knitr::kable(round(S,dig),booktabs = T, 
             caption = "Covariance Matrix of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position","center"))
```


```{r,message=FALSE,echo=F, results='hide',fig.show='hide'}

# correlation matrix
r <- cor(HCV_num,method = "pearson")

dig <- 4

# kable extra table creation
library(kableExtra)
knitr::kable(round(r,dig),booktabs = T, 
             caption = "Correlation Matrix of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position"))

```


```{r,fig.align='center',fig.cap="Correlation Heatmap for Laboratory Variables",echo=F, results='hide',fig.show='hide'}
# melt the rounded corr mat for plotting
library(reshape2)
r[lower.tri(r)] <- NA
melted_cormat <- melt(round(r,2),na.rm = T)

library(ggplot2)
ggheatmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "dodgerblue", high = "salmon", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson\nCorrelation") +
  theme_minimal()+ # minimal theme
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed() + 
geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  legend.justification = c(1, 0),
  legend.position = c(0.6, 0.8),
  legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                title.position = "top", title.hjust = 0.5))
# Print the heatmap
print(ggheatmap)

```



**Question 1d)**  
  
Make a scatterplot matrix of laboratory variables by labeling the points using different colors/symbols for different categories (add a legend). In the diagonal cells of this matrix, add QQ plot of each variable. Comment on the pairwise relationships between variables based on the scatterplots and heatmap. Which pairs of variables are useful in distinguishing between different categories and how?  
  
  

**Question 1d Answer)**   



The Variables PROT and ALB have the strongest correlation and is positive. These two variables also seem to have a a good linear relationship. The variable CREA seems to have many large outliers that may hide the true correlation and pairwise relationships to the other variables. The variable GGT has a fairly strong correlation and linear relationship with the variables ALP and AST. The strongest negative correlation is between BIL and CHE, but from the scatter plot this may be do to the outliers in category 3 (Cirrhosis). Lastly the last two pairwise relationship that jumps out to me is CHOL and CHE with a positive correlation of 0.43.   
   
The variables CHE and BIL look like they may be good to distinguish between variables because they both seem to be roughly separated. AST may be another variable that could be used to differentiate the categories because the groups linear regression lines are not parallel.   



**Question 1e)**  
  
For each laboratory variable, make side-by-side box plots for each category. Put the plots for all attribute variables in one page. Comment on the marginal distributions based on the different levels of the category variable.
  
  
**Question 1e Answer)**     
```{r,out.width="116%",fig.height= 15,fig.width=20,fig.align='center',fig.cap="Laboratory Box plots for each Category",message = FALSE,echo=F, results='hide',fig.show='hide'}
library(reshape2)

textsize <- 23
#HCV$Category <- factor(HCV$Category) 
long <- melt(HCV[,-c(2:3)],id.vars= "Category")
levels(long$Category) <- c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis')
ggplot(long, aes(x = variable, y = value, group = Category,fill = Category)) +
  geom_boxplot() +
  theme(axis.text.x =  element_blank()) +
  facet_wrap( ~ variable, scales = "free")+
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title.x = element_blank(), #change font size of axis text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title   
 
```

    
For the Variable ALB Category 0 (Blood donor) the marginal distribution looks to be symmetrix with many outliers, Category 1 (Hepatitis) the marginal distribution looks to be skewed to the right, category 2 (Fibrosis) the marginal distribution looks to be symmetric, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be slightly skewed the left.    
     
For the Variable ALP Category 0 (Blood donor) the marginal distribution looks to be roughly symmetric but there seems to be outliers to the right indicating a longer right tail, Category 1 (Hepatitis) the marginal distribution looks to be roughly symmetric with bigger outlier to the right which may indicate left skewness, category 2 (Fibrosis) the marginal distribution looks to be slightly skewed to the left because the median is to the right, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be roughly symmetric but has large outliers to the right.    
   
For the Variable ALT Category 0 (Blood donor) the marginal distribution looks to be skewed to the left because there are multiple outliers to the left, Category 1 (Hepatitis) the marginal distribution looks to be heavily left skewed because the median is close the lower quantile, category 2 (Fibrosis) the marginal distribution looks to be skewed to left in a lesser degree, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be skewed to the left and has many outliers to the right tail.  
   
For the Variable AST Category 0 (Blood donor) the marginal distribution looks to be roughly symmetric but has many outliers to the right that may indicate a longer tail to the right, Category 1 (Hepatitis) the marginal distribution looks to be skewed to the left because the median is close to the lower quantile, category 2 (Fibrosis) the marginal distribution looks to be the same but to lesser degree compared to Category 1, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be symmetric but has a longer right tail.    
   
For the Variable BIL Category 0 (Blood donor) the marginal distribution looks to have a long right tail because there are many outliers to the right, Category 1 (Hepatitis) the marginal distribution looks to be roughly symmetric but has two outliers giving a longer tail to the right, category 2 (Fibrosis) the marginal distribution looks to be skewed to the right because the median is to the left and there is an outlier to the right tail, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be skewed to the left because the median is to the left and there are two large outliers on the right side.     
  
For the Variable CHE Category 0 (Blood donor) the marginal distribution looks to be symmetric with a couple outliers on the right side, Category 1 (Hepatitis) the marginal distribution looks to be heavily skewed to the right, category 2 (Fibrosis) the marginal distribution looks to be skewed to the right but to a lesser degree in comparison with Category 1, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be skewed to the left with the median fairly close to the lower quantile.    
   
For the Variable CHOL Category 0 (Blood donor) the marginal distribution looks to be symmetric but has a large spread and many outliers to the right, Category 1 (Hepatitis) the marginal distribution looks to be symmetric and has one large outlier to the right, category 2 (Fibrosis) the marginal distribution looks to have a heavier tail to the right, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be slightly skewed to the left with outliers on both sides.    
   
For the Variable CREA Category 0 (Blood donor) the marginal distribution looks to be slightly skewed to the left with outlier to both sides, Category 1 (Hepatitis) the marginal distribution looks to be symmetric with an outlier to the right, category 2 (Fibrosis) the marginal distribution looks to have an outlier to the right , and lastly Category 3 (Cirrhosis) the marginal distribution looks to be hevily skewed to the left and has an enormous outlier to the right.    
   
For the Variable GGT Category 0 (Blood donor) the marginal distribution looks to be slightly skewed to the left and has a lot of outliers to the right side, Category 1 (Hepatitis) the marginal distribution looks to be left skewed, category 2 (Fibrosis) the marginal distribution looks to be symmetric, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be slightly skewed to the left with large outliers to the right.    
    
For the Variable PROT Category 0 (Blood donor) the marginal distribution looks to be symmetric but has multiple outliers to the left, Category 1 (Hepatitis) the marginal distribution looks to be slightly skewed to the left, category 2 (Fibrosis) the marginal distribution looks to be symmetric, and lastly Category 3 (Cirrhosis) the marginal distribution looks to be slightly skewed to the right.    
    
    
    
**Question 1f)**  
     
Create a panel plot consisting of the following three plots:       
\begin{enumerate}[label=(\roman*), nosep]
\item Scatterplot of CHE vs. CHOL with different colors/symbols for males and females. Add marginal plots for CHE and CHOL. Also, add circles to the same plot based on the Age variable (larger circles for larger ages). Carefully choose symbol sizes and line widths to minimize the crowded look.  
\item Box plot of Age grouped by Category. Label outliers by the value of Sex.    
\item Histogram of Age for category 1 superimposed by histogram of Age for category 3 (choose colors carefully to ensure that both histograms are visible).  
\end{enumerate}
Note: Divide the plotting area into three parts in such a manner that none of the plots get squeezed/distorted. Make efficient use of the plotting area in order to minimize white space. Comment on any pattern observed in the plots. 

**Question 1f Answer)**   

```{r,echo=F, results='hide',fig.show='hide'}
# library
library(ggplot2)
library(ggExtra)



p1 <- ggplot(HCV, aes(x=CHE, y=CHOL, color=Sex, size=Age,alpha = I(0.3))) +
      geom_point() +
      theme(legend.position="bottom")+ 
      scale_size(range = c(.5,6))+
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
#p1<- addSmallLegend(p1)
p11 <- ggMarginal(p1, type="histogram", fill = "slateblue")
#p11


```

```{r,out.height="80%",echo=F, results='hide',fig.show='hide'}

outlier <- function(x,var,group,label,coef = 1.5){
  x$outlier <- NA
  id <- c()
  for(i in 1:length(levels(x[,group]))){
    id <- c(id,which(x[x[,group] == levels(x[,group])[i],var] < (quantile(x[x[,group] == levels(x[,group])[i],var],0.25)[[1]] - coef*IQR(x[x[,group] == levels(x[,group])[i],var])) | x[x[,group] == levels(x[,group])[i],var] > (quantile(x[x[,group] == levels(x[,group])[i],var],0.75)[[1]] + coef*IQR(x[x[,group] == levels(x[,group])[i],var]))))
  }
  
  x[id,ncol(x)] <- x[id,label]
  return(x)
}

tmp2 <- outlier(tmp,var = 2,group = 1, label = 3)
p2 <- ggplot(tmp2, aes(x=Age,y=Category,fill = Category)) +
      geom_boxplot() + 
      geom_text(aes(label=outlier),na.rm=TRUE,nudge_y=0.2)+
      theme(legend.position="bottom") +
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
#p2
```

```{r,out.height="80%",echo=F, results='hide',fig.show='hide'}
HCVgraph <- HCV
levels(HCVgraph$Category) <- c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis')
p3 <- ggplot( HCVgraph[HCVgraph$Category == '1 = Hepatitis' | HCVgraph$Category =='3 = Cirrhosis',], aes(x=Age, fill=Category)) +
    geom_histogram( alpha=0.3, position = 'identity',binwidth = 8) +
    scale_fill_manual(values=c("dodgerblue", "salmon")) +
    labs(fill="")+
  theme(legend.position="bottom",
        text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
```


```{r,out.width="100%",fig.height= 15,fig.width=20,fig.align='center',fig.cap="3 different graphs",message = FALSE,echo=F, results='hide',fig.show='hide'}

grid.arrange(NULL, p11,p2,p3, nrow = 2)
```

  
There seems to be more males that are further away from the center in the plot between CHE and CHOL. 
There is also one outlier for age in category 0 that is a male. The histograms of age grouped by Category 1 and Category 3 are skewed to opposite sides indicating two possible different subpopulations. 

  
  
**Question 1g)**  
     
Make a 3D scatterplot of CHE, CHOL, and PROT. Use different colors to represent different categories. Make another version of the plot by adding vertical lines that connect the points to the floor. Comment on the relationship between the three variables and whether it varies by category.     
  
  
**Question 1g Answer)**     

```{r,out.width="70%",fig.align='center',fig.cap="3d Scatter plot of CHE, CHOL, and PROT grouped by Category",message = FALSE,echo=F, results='hide',fig.show='hide'}
library(scatterplot3d)
colors <- c(  "yellow2","dodgerblue","salmon","mediumspringgreen")
colors <- colors[as.numeric(HCV$Category)]
plot <- scatterplot3d(HCV[,c("CHE","CHOL","PROT")],pch = 16, color=colors,angle = 55)
legend(plot$xyz.convert(10, 13,40), col= c(  "yellow2","dodgerblue","salmon","mediumspringgreen"), bg="white", lty=c(1,1), lwd=2, yjust=0, legend =c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis'), cex = .8)
```

```{r,out.width="70%",fig.align='center',fig.cap="3d Scatter plot of CHE, CHOL, and PROT grouped by Category with vertical lines",message = FALSE,echo=F, results='hide',fig.show='hide'}
library(scatterplot3d)
colors <- c(  "yellow2","dodgerblue","salmon","mediumspringgreen")
colors <- colors[as.numeric(HCV$Category)]
plot <- scatterplot3d(HCV[,c("CHE","CHOL","PROT")],pch = 16, color=colors,angle = 55,type = "h")
legend(plot$xyz.convert(10, 13,40), col= c(  "yellow2","dodgerblue","salmon","mediumspringgreen"), bg="white", lty=c(1,1), lwd=2, yjust=0, legend =c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis'), cex = .8)
```

There Seems to be a clear grouping of Category 3 by itself in the 3d plot away from the other groups this may be mainly do to the CHE variable. It is harder to distinguish any other patterns from this view or angle.  
  
  
**Question 1h)**  
     
For this and the remaining parts, consider a subset of variables Age, ALB, CHE, CHOL, and PROT. Check univariate normality assumption (using all plots and measures discussed in class). If normality appears to be violated, explore transformations that may help. For each variable, include only one transformation and the corresponding plot (after transformation) that appears to be the most helpful.
  
**Question 1h Answer)**     
  
Transformations applied are as follows:    
AGE -> $AGE^{1/3}$   
ALB -> ALB (No transformation applied)   
CHE -> CHE (No transformation applied)   
CHOL -> $\sqrt{CHOL}$    
PROT -> $PROT^{4}$   
```{r,echo=F, results='hide',fig.show='hide'}
df <- HCV[,c("Age", "ALB", "CHE", "CHOL", "PROT")]
transdf <- df
```


```{r,fig.cap = "Transformation of $Age^{1/3}$ applied", out.width="50%",fig.align='center',results="hide",echo=F, results='hide',fig.show='hide'}
#ggplot(df, aes(x=Age)) +
#  geom_histogram(position="identity", fill="#69b3a2", color="#e9ecef", alpha=0.9)

#shapiro.test(df$Age)
## Normal Probability plot(q-q plot)
par(mfrow = c(1,2))
library(MASS)
library(car)
#qqPlot(df$Age,xlab = "Normal Quantiles",ylab="Age") + title("QQ Plot Age", cex.main = 1)
hist(df$Age,xlab = "Age",main="Histogram of Age",col = "dodgerblue")

#powerTransform(df$Age)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(df$Age,0.3027923), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(df$Age,0.3027923))
#shapiro.test(bcPower(df$Age,0.3027923))


#par(mfrow = c(1,2))
transdf$Age <- (df$Age)^(1/3)
#qqPlot(transdf$Age,xlab = "Normal Quantiles",ylab="Age^(1/3)") + title("QQ Plot", cex.main = 1)
hist(transdf$Age,xlab = "Age^(1/3)",main="Histogram of Age^(1/3)",col = "dodgerblue")

#x <- (df$Age)^(1/3)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="Age") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "Age",main="Histogram of Age",col = "dodgerblue")


```




```{r,fig.cap="No transformation applied for ALB",out.width="50%",fig.align='center',results="hide",echo=F, results='hide',fig.show='hide'}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$ALB,xlab = "Normal Quantiles",ylab="ALB") + title("QQ Plot ALB", cex.main = 1)
hist(df$ALB,xlab = "ALB",main="Histogram of ALB",col = "dodgerblue")






#powerTransform(df$ALB[-c(216,531)])

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(df$ALB,1.264823), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(df$ALB,1.264823))
#shapiro.test(bcPower(df$ALB,1.264823))

#x <- (df$ALB)^(2)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```



```{r,fig.cap="No Tranformation applied to CHE",out.width="50%",fig.align='center',results="hide",echo=F, results='hide',fig.show='hide'}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$CHE,xlab = "Normal Quantiles",ylab="CHE") + title("QQ Plot CHE", cex.main = 1)
hist(df$CHE,xlab = "CHE",main="Histogram of CHE",col = "dodgerblue")

#par(mfrow = c(1,2))
#transdf$CHE <- transdf$CHE^1.125288
#qqPlot(transdf$CHE,xlab = "Normal Quantiles",ylab="CHE^1.125288") + title("QQ Plot", cex.main = 1)
#hist(transdf$CHE,xlab = "CHE",main="Histogram of CHE^1.125288",col = "dodgerblue")



#y <- df$CHE

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,1.264823), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,1.264823))
#shapiro.test(bcPower(y,1.264823))

#x <- (df$CHE)^1.125288
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```



```{r,fig.cap="Transformation applied is $\\sqrt{CHOL}$",out.width="50%",fig.align='center',echo=F, results='hide',fig.show='hide'}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$CHOL,xlab = "Normal Quantiles",ylab="CHOL") + title("QQ Plot CHOL", cex.main = 1)
#hist(df$CHE,xlab = "CHOL",main="Histogram of CHOL",col = "dodgerblue")

transdf$CHOL <- sqrt(transdf$CHOL)#^0.5939165
qqPlot(transdf$CHOL,xlab = "Normal Quantiles",ylab="sqrt(CHOL)") + title("QQ Plot Sqrt(CHOL)", cex.main = 1)
#hist(transdf$CHOL,xlab = "sqrt(CHOL)",main="Histogram of sqrt(CHOL)",col = "dodgerblue")



#y <- df$CHOL

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,0.5939165), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,0.5939165))
#shapiro.test(bcPower(y,0.5939165))

#x <- sqrt(df$CHOL)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```


```{r,fig.cap="Transformation applied is $PROT^4$",out.width="50%",fig.align='center',results="hide",echo=F, results='hide',fig.show='hide'}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$PROT,xlab = "Normal Quantiles",ylab="PROT") + title("QQ Plot PROT", cex.main = 1)
#hist(df$PROT,xlab = "PROT",main="Histogram of PROT",col = "dodgerblue")

transdf$PROT <- (transdf$PROT)^4
qqPlot(transdf$PROT,xlab = "Normal Quantiles",ylab="PROT^4") + title("QQ Plot PROT^4", cex.main = 1)
#hist(transdf$CHOL,xlab = "sqrt(CHOL)",main="Histogram of sqrt(CHOL)",col = "dodgerblue")



#y <- df$PROT

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,3.722231), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,3.722231))
#shapiro.test(bcPower(y,3.722231))

#x <- (df$PROT)^4
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")

#x <- bcPower(y,3.722231)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```





**Question 1i)**  
     
Retain the transformations found above and check the multivariate normality assumption and find univariate and multivariate outliers. Comment about the normality assumption based on previous and this part.  
  
**Question 1i Answer)**     

The Chisq plot is not straight so this indicates that the multivariate normality assumption is violated. 
   
The Observations that are multivariate Outliers : 
43 183 209 216 259 287 315 317 344 403 529 531 532 533 545 553 571 574 575 576 581 584 588 589  
    
There are 0 univariate outliers for $Age^{1/3}$.
There are 27 univariate outliers for ALB.  
There are 23 univariate outliers for CHE.  
There are 10 univariate outliers for $\sqrt{CHOL}$.  
There are 19 univariate outliers for $PROT^4$.  

```{r,echo=F, results='hide',fig.show='hide'}
#Function for calculating chi-square plot (copy and paste)
chisplot <- function(x) {
    if (!is.matrix(x)) stop("x is not a matrix")

    ### determine dimensions
    n <- nrow(x)
    p <- ncol(x)
    #
    xbar <- apply(x, 2, mean)
    S <- var(x)
    S <- solve(S)
    index <- (1:n)/(n+1)
    #
    xcent <- t(t(x) - xbar)
    di <- apply(xcent, 1, function(x,S) x %*% S %*% x,S)
    
    print("Observations that are multivariate Outliers")
    print(which(di > qchisq(.99,p)))
    quant <- qchisq(index,p)
    plot(quant, sort(di), ylab = "Ordered distances",
         xlab = "Chi-square quantile", lwd=2,pch=1)
}

chisplot(as.matrix(scale(transdf)))
```




```{r,echo=F, results='hide',fig.show='hide'}
length(boxplot(transdf$Age)$out)
length(boxplot(transdf$ALB)$out)
length(boxplot(transdf$CHE)$out)
length(boxplot(transdf$CHOL)$out)
length(boxplot(transdf$PROT)$out)
```



```{r}

```





```{r}

```



```{r}

```



```{r}

```
\newpage
***Output***





```{r}
#HCV <- read.csv("https://utdallas.box.com/shared/static/6noqrx35iao10p4w15z7kx6egggiiq8f.csv")
HCV <- read.csv("Project_1_multivariate_analysis_and_assumptions/Data/HCV.csv")
HCV$Category <- factor(HCV$Category)
tmp <- HCV
HCV$Sex <- factor(HCV$Sex)
```
**Question 1**    



```{r}
HCV_num <- subset(HCV,select =-c(Category,Age,Sex))
#head(HCV_num)

```
  
**Question 1a)**    
  


```{r}
#Calculate Mahalanobis Distances
maha.dist<- sqrt(mahalanobis(HCV_num,center=colMeans(HCV_num),cov=cov(HCV_num)))
names(maha.dist) <- paste("Patient_",1:(dim(HCV_num)[1]),sep="")
#Sort the Distances
#sort(maha.dist)

# mahalanobis distance summary
maha.summary <- summary(maha.dist)
dig <- 4
table.maha.summary <- as.data.frame(
  cbind(round(maha.summary[[1]],dig), round(maha.summary[[2]],dig), round(maha.summary[[3]],dig), 
        round(maha.summary[[4]],dig), round(maha.summary[[5]],dig), round(maha.summary[[6]],dig)))
colnames(table.maha.summary) <- c("Min","1st.Quantile","Median","Mean","3rd.Quantile","Max")
table.maha.summary <- rbind(table.maha.summary,
                            c(names(which(maha.dist == min(maha.dist))), "", names(which(maha.dist == median(maha.dist))), "", "", names(which(maha.dist == max(maha.dist)))))

# kable extratable creation
library(kableExtra)
knitr::kable(table.maha.summary,booktabs = T, 
             caption = "Summary Statistics of Mahalanobis distances", 
             align = "cccccc") %>% 
  kable_styling(latex_options = c("hold_position"))
```



```{r}
# to look at patient 325 and 570
#HCV[c(325,570),]

library(gridExtra)

```


```{r}
# euclidean function
euclidean <- function(x) {
  dist <- sqrt(rowSums(x -  colMeans(x))^2)
  return(dist)
}

#Calculate Mahalanobis Distances
euclidean.dist <- euclidean(HCV_num)
names(euclidean.dist) <- paste("Patient_",1:(dim(HCV_num)[1]),sep="")
#Sort the Distances
#sort(euclidean.dist)

# euclidean distance summary
euclidean.summary <- summary(euclidean.dist)
dig <- 4
table.euclidean.summary <- as.data.frame(
  cbind(round(euclidean.summary[[1]],dig), round(euclidean.summary[[2]],dig), round(euclidean.summary[[3]],dig), 
        round(euclidean.summary[[4]],dig), round(euclidean.summary[[5]],dig), round(euclidean.summary[[6]],dig)))
colnames(table.euclidean.summary) <- c("Min","1st.Quantile","Median","Mean","3rd.Quantile","Max")
table.euclidean.summary <- rbind(table.euclidean.summary,
                            c(names(which(euclidean.dist == min(euclidean.dist))), "", names(which(euclidean.dist == median(euclidean.dist))), "", "", names(which(euclidean.dist == max(euclidean.dist)))))

# kable extratable creation
library(kableExtra)
knitr::kable(table.euclidean.summary,booktabs = T, 
             caption = "Summary Statistics of Euclidean distances", 
             align = "cccccc") %>% 
  kable_styling(latex_options = c("hold_position"))
```


   
**Question 1b)**     
   


```{r}
# covariance matrix

# X <- as.matrix(HCV_num)
# (t(X) %*% (diag(dim(X)[1]) - (1/dim(X)[1])*matrix(1,nrow = dim(X)[1],ncol = dim(X)[1])) %*% X)*(1/(dim(X)[1]-1))
S <- cov(HCV_num,method = "pearson")

# get two largest eigen pairs of of the ellipsoid
#eigen(S)
lam1 <- eigen(S)$values[1]
e1 <- eigen(S)$vectors[,1]
lam2 <- eigen(S)$values[2]
e2 <- eigen(S)$vectors[,2]

```



**Question 1c)**    
  

```{r}
# sample mean vector
xbar <- colMeans(HCV_num)

dig <- 4

# kable extratable creation
library(kableExtra)
knitr::kable(round(t(xbar),dig),booktabs = T, 
             caption = "Sample Mean Vector of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position"))
```


```{r}
# covariance matrix

# X <- as.matrix(HCV_num)
# (t(X) %*% (diag(dim(X)[1]) - (1/dim(X)[1])*matrix(1,nrow = dim(X)[1],ncol = dim(X)[1])) %*% X)*(1/(dim(X)[1]-1))
S <- cov(HCV_num,method = "pearson")



dig <- 3

# kable extra table creation
library(kableExtra)
knitr::kable(round(S,dig),booktabs = T, 
             caption = "Covariance Matrix of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position","center"))
```
\newpage

```{r,message=FALSE}

# correlation matrix
r <- cor(HCV_num,method = "pearson")

dig <- 4

# kable extra table creation
library(kableExtra)
knitr::kable(round(r,dig),booktabs = T, 
             caption = "Correlation Matrix of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position"))

```


```{r,fig.align='center',fig.cap="Correlation Heatmap for Laboratory Variables"}
# melt the rounded corr mat for plotting
library(reshape2)
r[lower.tri(r)] <- NA
melted_cormat <- melt(round(r,2),na.rm = T)

library(ggplot2)
ggheatmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "dodgerblue", high = "salmon", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson\nCorrelation") +
  theme_minimal()+ # minimal theme
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed() + 
geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  legend.justification = c(1, 0),
  legend.position = c(0.6, 0.8),
  legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                title.position = "top", title.hjust = 0.5))
# Print the heatmap
print(ggheatmap)

```



**Question 1d)**  
  




```{r,out.width="115%",fig.height= 20,fig.width=20,fig.align='left',fig.cap="Scatterplot Matrix with QQ-plot on the Diagonal for Laboratory Variables",message = FALSE,fig.subcap = "Grouped by Category"}

library(car)
# scatter plot matrix
scatterplotMatrix(~ ALB + ALP + ALT + AST + BIL + CHE + CHOL + CREA + GGT + PROT |Category,data = HCV, diagonal = list(method = "qqplot"),cex = .9,cex.axis = 3,cex.labels = 3,smooth = FALSE,legend = list(coords ="topleft",cex= 1.6,pt.cex = 1.6))
```

\newpage

**Question 1e)**  

```{r,out.width="116%",fig.height= 15,fig.width=20,fig.align='center',fig.cap="Laboratory Box plots for each Category",message = FALSE}
library(reshape2)

textsize <- 23
#HCV$Category <- factor(HCV$Category) 
long <- melt(HCV[,-c(2:3)],id.vars= "Category")
levels(long$Category) <- c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis')
ggplot(long, aes(x = variable, y = value, group = Category,fill = Category)) +
  geom_boxplot() +
  theme(axis.text.x =  element_blank()) +
  facet_wrap( ~ variable, scales = "free")+
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title.x = element_blank(), #change font size of axis text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title   
 
```


    
    
    
    
    
\newpage    
   
**Question 1f)**  
     



```{r}
# library
library(ggplot2)
library(ggExtra)



p1 <- ggplot(HCV, aes(x=CHE, y=CHOL, color=Sex, size=Age,alpha = I(0.3))) +
      geom_point() +
      theme(legend.position="bottom")+ 
      scale_size(range = c(.5,6))+
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
#p1<- addSmallLegend(p1)
p11 <- ggMarginal(p1, type="histogram", fill = "slateblue")
#p11


```

```{r,out.height="80%"}

outlier <- function(x,var,group,label,coef = 1.5){
  x$outlier <- NA
  id <- c()
  for(i in 1:length(levels(x[,group]))){
    id <- c(id,which(x[x[,group] == levels(x[,group])[i],var] < (quantile(x[x[,group] == levels(x[,group])[i],var],0.25)[[1]] - coef*IQR(x[x[,group] == levels(x[,group])[i],var])) | x[x[,group] == levels(x[,group])[i],var] > (quantile(x[x[,group] == levels(x[,group])[i],var],0.75)[[1]] + coef*IQR(x[x[,group] == levels(x[,group])[i],var]))))
  }
  
  x[id,ncol(x)] <- x[id,label]
  return(x)
}

tmp2 <- outlier(tmp,var = 2,group = 1, label = 3)
p2 <- ggplot(tmp2, aes(x=Age,y=Category,fill = Category)) +
      geom_boxplot() + 
      geom_text(aes(label=outlier),na.rm=TRUE,nudge_y=0.2)+
      theme(legend.position="bottom") +
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
#p2
```

```{r,out.height="80%"}
HCVgraph <- HCV
levels(HCVgraph$Category) <- c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis')
p3 <- ggplot( HCVgraph[HCVgraph$Category == '1 = Hepatitis' | HCVgraph$Category =='3 = Cirrhosis',], aes(x=Age, fill=Category)) +
    geom_histogram( alpha=0.3, position = 'identity',binwidth = 8) +
    scale_fill_manual(values=c("dodgerblue", "salmon")) +
    labs(fill="")+
  theme(legend.position="bottom",
        text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
```


```{r,out.width="100%",fig.height= 15,fig.width=20,fig.align='center',fig.cap="3 different graphs",message = FALSE}

grid.arrange(NULL, p11,p2,p3, nrow = 2)
```

  

  
\newpage 
**Question 1g)**  
     

```{r,out.width="70%",fig.align='center',fig.cap="3d Scatter plot of CHE, CHOL, and PROT grouped by Category",message = FALSE}
library(scatterplot3d)
colors <- c(  "yellow2","dodgerblue","salmon","mediumspringgreen")
colors <- colors[as.numeric(HCV$Category)]
plot <- scatterplot3d(HCV[,c("CHE","CHOL","PROT")],pch = 16, color=colors,angle = 55)
legend(plot$xyz.convert(10, 13,40), col= c(  "yellow2","dodgerblue","salmon","mediumspringgreen"), bg="white", lty=c(1,1), lwd=2, yjust=0, legend =c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis'), cex = .8)
```

```{r,out.width="70%",fig.align='center',fig.cap="3d Scatter plot of CHE, CHOL, and PROT grouped by Category with vertical lines",message = FALSE}
library(scatterplot3d)
colors <- c(  "yellow2","dodgerblue","salmon","mediumspringgreen")
colors <- colors[as.numeric(HCV$Category)]
plot <- scatterplot3d(HCV[,c("CHE","CHOL","PROT")],pch = 16, color=colors,angle = 55,type = "h")
legend(plot$xyz.convert(10, 13,40), col= c(  "yellow2","dodgerblue","salmon","mediumspringgreen"), bg="white", lty=c(1,1), lwd=2, yjust=0, legend =c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis'), cex = .8)
```


\newpage  
**Question 1h)**  
     

```{r}
df <- HCV[,c("Age", "ALB", "CHE", "CHOL", "PROT")]
transdf <- df
```


```{r,fig.cap = "Transformation of $Age^{1/3}$ applied", out.width="50%",fig.align='center',results="hide"}
#ggplot(df, aes(x=Age)) +
#  geom_histogram(position="identity", fill="#69b3a2", color="#e9ecef", alpha=0.9)

#shapiro.test(df$Age)
## Normal Probability plot(q-q plot)
par(mfrow = c(1,2))
library(MASS)
library(car)
#qqPlot(df$Age,xlab = "Normal Quantiles",ylab="Age") + title("QQ Plot Age", cex.main = 1)
hist(df$Age,xlab = "Age",main="Histogram of Age",col = "dodgerblue")

#powerTransform(df$Age)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(df$Age,0.3027923), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(df$Age,0.3027923))
#shapiro.test(bcPower(df$Age,0.3027923))


#par(mfrow = c(1,2))
transdf$Age <- (df$Age)^(1/3)
#qqPlot(transdf$Age,xlab = "Normal Quantiles",ylab="Age^(1/3)") + title("QQ Plot", cex.main = 1)
hist(transdf$Age,xlab = "Age^(1/3)",main="Histogram of Age^(1/3)",col = "dodgerblue")

#x <- (df$Age)^(1/3)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="Age") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "Age",main="Histogram of Age",col = "dodgerblue")


```




```{r,fig.cap="No transformation applied for ALB",out.width="50%",fig.align='center',results="hide"}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$ALB,xlab = "Normal Quantiles",ylab="ALB") + title("QQ Plot ALB", cex.main = 1)
hist(df$ALB,xlab = "ALB",main="Histogram of ALB",col = "dodgerblue")






#powerTransform(df$ALB[-c(216,531)])

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(df$ALB,1.264823), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(df$ALB,1.264823))
#shapiro.test(bcPower(df$ALB,1.264823))

#x <- (df$ALB)^(2)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```



```{r,fig.cap="No Tranformation applied to CHE",out.width="50%",fig.align='center',results="hide"}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$CHE,xlab = "Normal Quantiles",ylab="CHE") + title("QQ Plot CHE", cex.main = 1)
hist(df$CHE,xlab = "CHE",main="Histogram of CHE",col = "dodgerblue")

#par(mfrow = c(1,2))
#transdf$CHE <- transdf$CHE^1.125288
#qqPlot(transdf$CHE,xlab = "Normal Quantiles",ylab="CHE^1.125288") + title("QQ Plot", cex.main = 1)
#hist(transdf$CHE,xlab = "CHE",main="Histogram of CHE^1.125288",col = "dodgerblue")



#y <- df$CHE

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,1.264823), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,1.264823))
#shapiro.test(bcPower(y,1.264823))

#x <- (df$CHE)^1.125288
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```



```{r,fig.cap="Transformation applied is $\\sqrt{CHOL}$",out.width="50%",fig.align='center',results="hide"}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$CHOL,xlab = "Normal Quantiles",ylab="CHOL") + title("QQ Plot CHOL", cex.main = 1)
#hist(df$CHE,xlab = "CHOL",main="Histogram of CHOL",col = "dodgerblue")

transdf$CHOL <- sqrt(transdf$CHOL)#^0.5939165
qqPlot(transdf$CHOL,xlab = "Normal Quantiles",ylab="sqrt(CHOL)") + title("QQ Plot Sqrt(CHOL)", cex.main = 1)
#hist(transdf$CHOL,xlab = "sqrt(CHOL)",main="Histogram of sqrt(CHOL)",col = "dodgerblue")



#y <- df$CHOL

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,0.5939165), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,0.5939165))
#shapiro.test(bcPower(y,0.5939165))

#x <- sqrt(df$CHOL)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```


```{r,fig.cap="Transformation applied is $PROT^4$",out.width="50%",fig.align='center',results="hide"}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$PROT,xlab = "Normal Quantiles",ylab="PROT") + title("QQ Plot PROT", cex.main = 1)
#hist(df$PROT,xlab = "PROT",main="Histogram of PROT",col = "dodgerblue")

transdf$PROT <- (transdf$PROT)^4
qqPlot(transdf$PROT,xlab = "Normal Quantiles",ylab="PROT^4") + title("QQ Plot PROT^4", cex.main = 1)
#hist(transdf$CHOL,xlab = "sqrt(CHOL)",main="Histogram of sqrt(CHOL)",col = "dodgerblue")



#y <- df$PROT

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,3.722231), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,3.722231))
#shapiro.test(bcPower(y,3.722231))

#x <- (df$PROT)^4
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")

#x <- bcPower(y,3.722231)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```


\newpage


**Question 1i)**  
     

```{r}
#Function for calculating chi-square plot (copy and paste)
chisplot <- function(x) {
    if (!is.matrix(x)) stop("x is not a matrix")

    ### determine dimensions
    n <- nrow(x)
    p <- ncol(x)
    #
    xbar <- apply(x, 2, mean)
    S <- var(x)
    S <- solve(S)
    index <- (1:n)/(n+1)
    #
    xcent <- t(t(x) - xbar)
    di <- apply(xcent, 1, function(x,S) x %*% S %*% x,S)
    
    print("Observations that are multivariate Outliers")
    print(which(di > qchisq(.99,p)))
    quant <- qchisq(index,p)
    plot(quant, sort(di), ylab = "Ordered distances",
         xlab = "Chi-square quantile", lwd=2,pch=1)
}

chisplot(as.matrix(scale(transdf)))
```




```{r,fig.show='hide'}
length(boxplot(transdf$Age)$out)
length(boxplot(transdf$ALB)$out)
length(boxplot(transdf$CHE)$out)
length(boxplot(transdf$CHOL)$out)
length(boxplot(transdf$PROT)$out)
```






\newpage


***CODE***



```{r,echo=T, results='hide',fig.show='hide'}
#HCV <- read.csv("https://utdallas.box.com/shared/static/6noqrx35iao10p4w15z7kx6egggiiq8f.csv")
HCV <- read.csv("Project_1_multivariate_analysis_and_assumptions/Data/HCV.csv")
HCV$Category <- factor(HCV$Category)
tmp <- HCV
HCV$Sex <- factor(HCV$Sex)
```
**Question 1**    



```{r,echo=T, results='hide',fig.show='hide'}
HCV_num <- subset(HCV,select =-c(Category,Age,Sex))
#head(HCV_num)

```
  
**Question 1a)**    
  


```{r,echo=T, results='hide',fig.show='hide'}
#Calculate Mahalanobis Distances
maha.dist<- sqrt(mahalanobis(HCV_num,center=colMeans(HCV_num),cov=cov(HCV_num)))
names(maha.dist) <- paste("Patient_",1:(dim(HCV_num)[1]),sep="")
#Sort the Distances
#sort(maha.dist)

# mahalanobis distance summary
maha.summary <- summary(maha.dist)
dig <- 4
table.maha.summary <- as.data.frame(
  cbind(round(maha.summary[[1]],dig), round(maha.summary[[2]],dig), round(maha.summary[[3]],dig), 
        round(maha.summary[[4]],dig), round(maha.summary[[5]],dig), round(maha.summary[[6]],dig)))
colnames(table.maha.summary) <- c("Min","1st.Quantile","Median","Mean","3rd.Quantile","Max")
table.maha.summary <- rbind(table.maha.summary,
                            c(names(which(maha.dist == min(maha.dist))), "", names(which(maha.dist == median(maha.dist))), "", "", names(which(maha.dist == max(maha.dist)))))

# kable extratable creation
library(kableExtra)
knitr::kable(table.maha.summary,booktabs = T, 
             caption = "Summary Statistics of Mahalanobis distances", 
             align = "cccccc") %>% 
  kable_styling(latex_options = c("hold_position"))
```



```{r,echo=T, results='hide',fig.show='hide'}
# to look at patient 325 and 570
#HCV[c(325,570),]

library(gridExtra)

```


```{r,echo=T, results='hide',fig.show='hide'}
# euclidean function
euclidean <- function(x) {
  dist <- sqrt(rowSums(x -  colMeans(x))^2)
  return(dist)
}

#Calculate Mahalanobis Distances
euclidean.dist <- euclidean(HCV_num)
names(euclidean.dist) <- paste("Patient_",1:(dim(HCV_num)[1]),sep="")
#Sort the Distances
#sort(euclidean.dist)

# euclidean distance summary
euclidean.summary <- summary(euclidean.dist)
dig <- 4
table.euclidean.summary <- as.data.frame(
  cbind(round(euclidean.summary[[1]],dig), round(euclidean.summary[[2]],dig), round(euclidean.summary[[3]],dig), 
        round(euclidean.summary[[4]],dig), round(euclidean.summary[[5]],dig), round(euclidean.summary[[6]],dig)))
colnames(table.euclidean.summary) <- c("Min","1st.Quantile","Median","Mean","3rd.Quantile","Max")
table.euclidean.summary <- rbind(table.euclidean.summary,
                            c(names(which(euclidean.dist == min(euclidean.dist))), "", names(which(euclidean.dist == median(euclidean.dist))), "", "", names(which(euclidean.dist == max(euclidean.dist)))))

# kable extratable creation
library(kableExtra)
knitr::kable(table.euclidean.summary,booktabs = T, 
             caption = "Summary Statistics of Euclidean distances", 
             align = "cccccc") %>% 
  kable_styling(latex_options = c("hold_position"))
```


   
**Question 1b)**     
   


```{r,echo=T, results='hide',fig.show='hide'}
# covariance matrix

# X <- as.matrix(HCV_num)
# (t(X) %*% (diag(dim(X)[1]) - (1/dim(X)[1])*matrix(1,nrow = dim(X)[1],ncol = dim(X)[1])) %*% X)*(1/(dim(X)[1]-1))
S <- cov(HCV_num,method = "pearson")

# get two largest eigen pairs of of the ellipsoid
#eigen(S)
lam1 <- eigen(S)$values[1]
e1 <- eigen(S)$vectors[,1]
lam2 <- eigen(S)$values[2]
e2 <- eigen(S)$vectors[,2]

```



**Question 1c)**    
  

```{r,echo=T, results='hide',fig.show='hide'}
# sample mean vector
xbar <- colMeans(HCV_num)

dig <- 4

# kable extratable creation
library(kableExtra)
knitr::kable(round(t(xbar),dig),booktabs = T, 
             caption = "Sample Mean Vector of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position"))
```


```{r,echo=T, results='hide',fig.show='hide'}
# covariance matrix

# X <- as.matrix(HCV_num)
# (t(X) %*% (diag(dim(X)[1]) - (1/dim(X)[1])*matrix(1,nrow = dim(X)[1],ncol = dim(X)[1])) %*% X)*(1/(dim(X)[1]-1))
S <- cov(HCV_num,method = "pearson")



dig <- 3

# kable extra table creation
library(kableExtra)
knitr::kable(round(S,dig),booktabs = T, 
             caption = "Covariance Matrix of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position","center"))
```


```{r,echo=T, results='hide',fig.show='hide'}

# correlation matrix
r <- cor(HCV_num,method = "pearson")

dig <- 4

# kable extra table creation
library(kableExtra)
knitr::kable(round(r,dig),booktabs = T, 
             caption = "Correlation Matrix of the Laboratory variables") %>% 
  kable_styling(latex_options = c("hold_position"))

```


```{r,echo=T, results='hide',fig.show='hide'}
# melt the rounded corr mat for plotting
library(reshape2)
r[lower.tri(r)] <- NA
melted_cormat <- melt(round(r,2),na.rm = T)

library(ggplot2)
ggheatmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "dodgerblue", high = "salmon", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson\nCorrelation") +
  theme_minimal()+ # minimal theme
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed() + 
geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) +
theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  legend.justification = c(1, 0),
  legend.position = c(0.6, 0.8),
  legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                title.position = "top", title.hjust = 0.5))
# Print the heatmap
print(ggheatmap)

```



**Question 1d)**  
  




```{r,echo=T, results='hide',fig.show='hide'}

library(car)
# scatter plot matrix
scatterplotMatrix(~ ALB + ALP + ALT + AST + BIL + CHE + CHOL + CREA + GGT + PROT |Category,data = HCV, diagonal = list(method = "qqplot"),cex = .9,cex.axis = 3,cex.labels = 3,smooth = FALSE,legend = list(coords ="topleft",cex= 1.6,pt.cex = 1.6))
```



**Question 1e)**  

```{r,echo=T, results='hide',fig.show='hide'}
library(reshape2)

textsize <- 23
#HCV$Category <- factor(HCV$Category) 
long <- melt(HCV[,-c(2:3)],id.vars= "Category")
levels(long$Category) <- c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis')
ggplot(long, aes(x = variable, y = value, group = Category,fill = Category)) +
  geom_boxplot() +
  theme(axis.text.x =  element_blank()) +
  facet_wrap( ~ variable, scales = "free")+
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title.x = element_blank(), #change font size of axis text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title   
 
```


    
    
    
    
    
    
   
**Question 1f)**  
     



```{r,echo=T, results='hide',fig.show='hide'}
# library
library(ggplot2)
library(ggExtra)



p1 <- ggplot(HCV, aes(x=CHE, y=CHOL, color=Sex, size=Age,alpha = I(0.3))) +
      geom_point() +
      theme(legend.position="bottom")+ 
      scale_size(range = c(.5,6))+
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
#p1<- addSmallLegend(p1)
p11 <- ggMarginal(p1, type="histogram", fill = "slateblue")
#p11


```

```{r,out.height="80%",echo=T, results='hide',fig.show='hide'}

outlier <- function(x,var,group,label,coef = 1.5){
  x$outlier <- NA
  id <- c()
  for(i in 1:length(levels(x[,group]))){
    id <- c(id,which(x[x[,group] == levels(x[,group])[i],var] < (quantile(x[x[,group] == levels(x[,group])[i],var],0.25)[[1]] - coef*IQR(x[x[,group] == levels(x[,group])[i],var])) | x[x[,group] == levels(x[,group])[i],var] > (quantile(x[x[,group] == levels(x[,group])[i],var],0.75)[[1]] + coef*IQR(x[x[,group] == levels(x[,group])[i],var]))))
  }
  
  x[id,ncol(x)] <- x[id,label]
  return(x)
}

tmp2 <- outlier(tmp,var = 2,group = 1, label = 3)
p2 <- ggplot(tmp2, aes(x=Age,y=Category,fill = Category)) +
      geom_boxplot() + 
      geom_text(aes(label=outlier),na.rm=TRUE,nudge_y=0.2)+
      theme(legend.position="bottom") +
  theme(text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
#p2
```

```{r,out.height="80%",echo=T, results='hide',fig.show='hide'}
HCVgraph <- HCV
levels(HCVgraph$Category) <- c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis')
p3 <- ggplot( HCVgraph[HCVgraph$Category == '1 = Hepatitis' | HCVgraph$Category =='3 = Cirrhosis',], aes(x=Age, fill=Category)) +
    geom_histogram( alpha=0.3, position = 'identity',binwidth = 8) +
    scale_fill_manual(values=c("dodgerblue", "salmon")) +
    labs(fill="")+
  theme(legend.position="bottom",
        text=element_text(size=textsize), #change font size of all text
        axis.title=element_text(size=textsize), #change font size of axis titles
        plot.title=element_text(size=textsize), #change font size of plot title
        legend.text=element_text(size=textsize), #change font size of legend text
        legend.title=element_text(size=textsize))#change font size of legend title  
```


```{r,echo=T, results='hide',fig.show='hide'}

grid.arrange(NULL, p11,p2,p3, nrow = 2)
```

  

  
  
**Question 1g)**  
     

```{r,echo=T, results='hide',fig.show='hide'}
library(scatterplot3d)
colors <- c(  "yellow2","dodgerblue","salmon","mediumspringgreen")
colors <- colors[as.numeric(HCV$Category)]
plot <- scatterplot3d(HCV[,c("CHE","CHOL","PROT")],pch = 16, color=colors,angle = 55)
legend(plot$xyz.convert(10, 13,40), col= c(  "yellow2","dodgerblue","salmon","mediumspringgreen"), bg="white", lty=c(1,1), lwd=2, yjust=0, legend =c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis'), cex = .8)
```

```{r,echo=T, results='hide',fig.show='hide'}
library(scatterplot3d)
colors <- c(  "yellow2","dodgerblue","salmon","mediumspringgreen")
colors <- colors[as.numeric(HCV$Category)]
plot <- scatterplot3d(HCV[,c("CHE","CHOL","PROT")],pch = 16, color=colors,angle = 55,type = "h")
legend(plot$xyz.convert(10, 13,40), col= c(  "yellow2","dodgerblue","salmon","mediumspringgreen"), bg="white", lty=c(1,1), lwd=2, yjust=0, legend =c('0 = Blood Donor', '1 = Hepatitis', '2 = Fibrosis','3 = Cirrhosis'), cex = .8)
```


  
**Question 1h)**  
     

```{r,echo=T, results='hide',fig.show='hide'}
df <- HCV[,c("Age", "ALB", "CHE", "CHOL", "PROT")]
transdf <- df
```


```{r,echo=T, results='hide',fig.show='hide'}
#ggplot(df, aes(x=Age)) +
#  geom_histogram(position="identity", fill="#69b3a2", color="#e9ecef", alpha=0.9)

#shapiro.test(df$Age)
## Normal Probability plot(q-q plot)
par(mfrow = c(1,2))
library(MASS)
library(car)
#qqPlot(df$Age,xlab = "Normal Quantiles",ylab="Age") + title("QQ Plot Age", cex.main = 1)
hist(df$Age,xlab = "Age",main="Histogram of Age",col = "dodgerblue")

#powerTransform(df$Age)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(df$Age,0.3027923), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(df$Age,0.3027923))
#shapiro.test(bcPower(df$Age,0.3027923))


#par(mfrow = c(1,2))
transdf$Age <- (df$Age)^(1/3)
#qqPlot(transdf$Age,xlab = "Normal Quantiles",ylab="Age^(1/3)") + title("QQ Plot", cex.main = 1)
hist(transdf$Age,xlab = "Age^(1/3)",main="Histogram of Age^(1/3)",col = "dodgerblue")

#x <- (df$Age)^(1/3)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="Age") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "Age",main="Histogram of Age",col = "dodgerblue")


```




```{r,echo=T, results='hide',fig.show='hide'}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$ALB,xlab = "Normal Quantiles",ylab="ALB") + title("QQ Plot ALB", cex.main = 1)
hist(df$ALB,xlab = "ALB",main="Histogram of ALB",col = "dodgerblue")






#powerTransform(df$ALB[-c(216,531)])

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(df$ALB,1.264823), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(df$ALB,1.264823))
#shapiro.test(bcPower(df$ALB,1.264823))

#x <- (df$ALB)^(2)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```



```{r,echo=T, results='hide',fig.show='hide'}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$CHE,xlab = "Normal Quantiles",ylab="CHE") + title("QQ Plot CHE", cex.main = 1)
hist(df$CHE,xlab = "CHE",main="Histogram of CHE",col = "dodgerblue")

#par(mfrow = c(1,2))
#transdf$CHE <- transdf$CHE^1.125288
#qqPlot(transdf$CHE,xlab = "Normal Quantiles",ylab="CHE^1.125288") + title("QQ Plot", cex.main = 1)
#hist(transdf$CHE,xlab = "CHE",main="Histogram of CHE^1.125288",col = "dodgerblue")



#y <- df$CHE

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,1.264823), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,1.264823))
#shapiro.test(bcPower(y,1.264823))

#x <- (df$CHE)^1.125288
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```



```{r,echo=T, results='hide',fig.show='hide'}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$CHOL,xlab = "Normal Quantiles",ylab="CHOL") + title("QQ Plot CHOL", cex.main = 1)
#hist(df$CHE,xlab = "CHOL",main="Histogram of CHOL",col = "dodgerblue")

transdf$CHOL <- sqrt(transdf$CHOL)#^0.5939165
qqPlot(transdf$CHOL,xlab = "Normal Quantiles",ylab="sqrt(CHOL)") + title("QQ Plot Sqrt(CHOL)", cex.main = 1)
#hist(transdf$CHOL,xlab = "sqrt(CHOL)",main="Histogram of sqrt(CHOL)",col = "dodgerblue")



#y <- df$CHOL

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,0.5939165), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,0.5939165))
#shapiro.test(bcPower(y,0.5939165))

#x <- sqrt(df$CHOL)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```


```{r,echo=T, results='hide',fig.show='hide'}
#par(mfrow = c(2,2))
par(mfrow = c(1,2))
library(MASS)
library(car)
qqPlot(df$PROT,xlab = "Normal Quantiles",ylab="PROT") + title("QQ Plot PROT", cex.main = 1)
#hist(df$PROT,xlab = "PROT",main="Histogram of PROT",col = "dodgerblue")

transdf$PROT <- (transdf$PROT)^4
qqPlot(transdf$PROT,xlab = "Normal Quantiles",ylab="PROT^4") + title("QQ Plot PROT^4", cex.main = 1)
#hist(transdf$CHOL,xlab = "sqrt(CHOL)",main="Histogram of sqrt(CHOL)",col = "dodgerblue")



#y <- df$PROT

#powerTransform(y)

#QQ Plot of the Box-Cox transformed data with the specified power
#qqnorm(bcPower(y,3.722231), main="QQ Plot after Box-Cox Transformation")
#qqline(bcPower(y,3.722231))
#shapiro.test(bcPower(y,3.722231))

#x <- (df$PROT)^4
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")

#x <- bcPower(y,3.722231)
#shapiro.test(x)
## Normal Probability plot(q-q plot)
#par(mfrow = c(1,2))
#library(MASS)
#library(car)
#qqPlot(x,xlab = "Normal Quantiles",ylab="x") + title("QQ Plot", cex.main = 1)
#hist(x,xlab = "x",main="Histogram of x",col = "dodgerblue")
```





**Question 1i)**  
     

```{r,echo=T, results='hide',fig.show='hide'}
#Function for calculating chi-square plot (copy and paste)
chisplot <- function(x) {
    if (!is.matrix(x)) stop("x is not a matrix")

    ### determine dimensions
    n <- nrow(x)
    p <- ncol(x)
    #
    xbar <- apply(x, 2, mean)
    S <- var(x)
    S <- solve(S)
    index <- (1:n)/(n+1)
    #
    xcent <- t(t(x) - xbar)
    di <- apply(xcent, 1, function(x,S) x %*% S %*% x,S)
    
    print("Observations that are multivariate Outliers")
    print(which(di > qchisq(.99,p)))
    quant <- qchisq(index,p)
    plot(quant, sort(di), ylab = "Ordered distances",
         xlab = "Chi-square quantile", lwd=2,pch=1)
}

chisplot(as.matrix(scale(transdf)))
```




```{r,echo=T, results='hide',fig.show='hide'}
length(boxplot(transdf$Age)$out)
length(boxplot(transdf$ALB)$out)
length(boxplot(transdf$CHE)$out)
length(boxplot(transdf$CHOL)$out)
length(boxplot(transdf$PROT)$out)
```











